Finally tagless, partially evaluated

### 2007-11-24 tagless/talk.tex
Some small fixes to be consistent about ML notation

    \bigskip
    \begin{tikzpicture}
        \draw node [lang] (lang0) {contract};
        \foreach \label/\prev/\cur in {grammar/0/1,music/1/2,$\lambda$-term/2/3}
            \draw (lang\prev.east)+(.3,0) node [lang] (lang\cur) {\label};
        \draw (lang3.east)+(.2,0) node [anchor=west] {\dots};
    \onslide<2>
        \draw (lang0)+(-110:2cm)   node (terp01) {value};
        \draw (lang0)+( -85:2.5cm) node (terp02) {schedule};
        \draw (lang1)+(-110:2cm)   node (terp11) {parse};
        \draw (lang1)+( -85:2.5cm) node (terp12) {pretty-print};
        \draw (lang2)+(-110:2cm)   node (terp21) {typeset};
        \draw (lang2)+( -85:2.5cm) node (terp22) {perform};
        \draw (lang3)+(-110:2cm)   node (terp31) {evaluate};
        \draw (lang3)+( -85:2.5cm) node (terp32) {compile};
        \foreach \lang in {0,...,3} {
            \draw (lang\lang)+(-65:1.5cm) node {\dots};
            \foreach \terp in {1,...,2}
                \draw [->] (lang\lang) -- (terp\lang\terp);
        }
    \onslide<3>
        \draw (lang3)+(-148:4.5cm)  node (terp31) {evaluate};
        \draw (lang3)+(-133:4.25cm) node (terp32) {compile};
        \draw (lang3)+(-113:4cm)    node (terp33) {specialize};
        \draw (lang3)+( -93:4.4cm)  node (terp34) {CPS transform};
        \draw (lang3)+( -73:4cm)    node (terp35) {pretty-print};
        \draw (lang3)+( -58:3cm)    node {\dots};
        \foreach \terp in {1,...,5} \draw [->] (lang3) -- (terp3\terp);
    \end{tikzpicture}

@@

\begin{frame}{How to preserve types?}
    \def\bool{\onslide<5->{ \rlap{\alert{bool}}}}
    \begin{tikzpicture}
        \draw node [lang] (lang)
          {\texttt{\alt<6->{\onslide<6>{\alert{leq} (\alert{lit} 3, }%
                            \only<7>{\makebox[0pt]{\sffamily
                                      Simply typed $\lambda$-calculus}}%
                            \onslide<6>{\alert{lit} 4):}}
                           {\onslide<2->{Leq (Lit 3, }%
                            \only<1>{\makebox[0pt]{\sffamily
                                      \alert{Typed} source language}}%
                            \onslide<2->{Lit 4):}}\
                   \only<6>{\rlap{\alert{bool}}}%
                   \onslide<2-5>{term\bool}}};
        \draw [->] (lang) --
            node [above,midway,anchor=east]
              {\only<1>{\alert{Typed} metalanguage\hspace*{2em}}%
               \only<2>{evaluate}%
               \only<7>{\begin{tabular}{@{}r@{\qquad}}
                        Haskell constructor classes\\
                        or ML functors
                        \end{tabular}}}
            node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
                {\only<-2,7>{\color{bg}}\ttfamily
                \alt<6->{\only<6>{\alert<6>{lit:\ int -> int}\\
                lit i = i\\[1ex]
                \alert<6>{leq:\ (int,int) -> bool}\\
                leq (i,j) = i <= j}}
                {\only<2-5>{\alert<5>{eval:\ term\only<5>{ 'a} -> value\only<5>{ 'a}}\\
                eval (Lit i) = Int i\\
                eval (Leq (m,n)) =\\
                \quad \alert<4>{match (eval m, eval n) with\\
                \quad (Int i, Int j) ->} Bool (i <= j)}}}
            +(-100:4cm)
            node [anchor=north,text depth=0pt,text height=9pt]
              {\texttt{\llap{\onslide<2-5>{Bool}
                             \onslide<2-6>{true:}%
                             \only<1>{\rlap{\sffamily
                                            \alert{Typed} target languages}}%
                             \only<7>{\rlap{\sffamily evaluate}}}\
                       \only<6>{\rlap{\alert{bool}}}%
                       \onslide<2-5>{value\bool}}};
        \draw [->] (lang) -- +(-80:3.5cm)
            node [anchor=north] {\only<7>{compile}};
        \draw [->] (lang) -- +(-65:3cm)
            node [anchor=north west,inner xsep=0pt] {\only<7>{specialize}};
        \draw [->] (lang) -- +(-50:2.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<7>{CPS transform}};
        \draw      (lang)    +(-35:2cm) node {\dots};
    \end{tikzpicture}

@@

    \strut
    \only<-3>{It should be obvious in the metalanguage that interpreting
    a well-typed source term yields a well-typed target term.}%
    \only<4>{The term should be well-typed (and closed), so pattern matching
    in the metalanguage should always \textbf{obviously} succeed.}%
    \only<5>{Previous solutions use (and motivate) fancier types:\\
    GADT, dependent types.}%
    \only<6>{Our simple solution is to be \textbf{finally tagless}:\\
    replace term constructors by cogen functions.}%
    \only<7>{To accommodate multiple interpretations,\\
    the term abstracts over the cogen functions and their types.}%
    \strut
\end{frame}

>>>>

    \bigskip
    \begin{tikzpicture}
        \draw node [lang] (lang0) {contract};
        \foreach \label/\prev/\cur in {grammar/0/1,music/1/2,$\lambda$-term/2/3}
            \draw (lang\prev.east)+(.3,0) node [lang] (lang\cur) {\label};
        \draw (lang3.east)+(.2,0) node [anchor=west] {\dots};
    \onslide<2>
        \draw (lang0)+(-110:2cm)   node (terp01) {value};
        \draw (lang0)+( -85:2.5cm) node (terp02) {schedule};
        \draw (lang1)+(-110:2cm)   node (terp11) {parse};
        \draw (lang1)+( -85:2.5cm) node (terp12) {pretty-print};
        \draw (lang2)+(-110:2cm)   node (terp21) {typeset};
        \draw (lang2)+( -85:2.5cm) node (terp22) {perform};
        \draw (lang3)+(-110:2cm)   node (terp31) {evaluate};
        \draw (lang3)+( -85:2.5cm) node (terp32) {compile};
        \foreach \lang in {0,...,3} {
            \draw (lang\lang)+(-65:1.5cm) node {\dots};
            \foreach \terp in {1,...,2}
                \draw [->] (lang\lang) -- (terp\lang\terp);
        }
    \onslide<3>
        \draw (lang3)+(-148:4.5cm)  node [inner xsep=0pt] (terp31) {evaluate};
        \draw (lang3)+(-133:4.25cm) node [inner xsep=0pt] (terp32) {compile};
        \draw (lang3)+(-113:4cm)    node [inner xsep=0pt] (terp33) {specialize};
        \draw (lang3)+( -93:4.4cm)  node [inner xsep=0pt] (terp34) {CPS transform};
        \draw (lang3)+( -73:4cm)    node [inner xsep=0pt] (terp35) {pretty-print};
        \draw (lang3)+( -58:3cm)    node {\dots};
        \foreach \terp in {1,...,5} \draw [->] (lang3) -- (terp3\terp);
    \end{tikzpicture}

@@

\begin{frame}{Simple type preservation}
    \def\bool{\only<5-6>{\ \alert{bool}}}
    \begin{tikzpicture}
        \draw node [lang] (lang)
          {\texttt{\alt<6->{\onslide<6>{\alert{leq} (\alert{lit} 3, }%
                            \only<7>{\makebox[0pt]{\sffamily
                                      Simply typed $\lambda$-calculus}}%
                            \onslide<6>{\alert{lit} 4):}}
                           {\onslide<2->{LEQ (LIT 3, }%
                            \only<1>{\makebox[0pt]{\sffamily
                                      \alert{Typed} source language}}%
                            \onslide<2->{LIT 4):}}%
                   \makebox[3em][l]{\bool\only<2-5>{\ term}}}};
        \draw [->] (lang) --
            node [above,midway,anchor=east]
              {\only<1>{\alert{Typed} metalanguage\hspace*{2em}}%
               \only<2>{evaluate}%
               \only<7>{\begin{tabular}{@{}r@{\qquad}}
                        Haskell constructor classes\\
                        or ML functors
                        \end{tabular}}}
            node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
                {\only<-2,7>{\color{bg}}\ttfamily
                \alt<6->{\only<6>{\alert<6>{lit:\ int -> int}\\
                lit i = i\\[1ex]
                \alert<6>{leq:\ int * int -> bool}\\
                leq (i,j) = i <= j}}
                {\only<2-5>{\alert<5>{eval:\ \only<5>{'a }term -> \only<5>{'a }value}\\
                eval (LIT i) = INT i\\
                eval (LEQ (m,n)) =\\
                \quad \alert<4>{match (eval m, eval n) with\\
                \quad (INT i, INT j) ->} BOOL (i <= j)}}}
            +(-100:4cm)
            node [anchor=north,text depth=0pt,text height=9pt]
              {\texttt{\llap{\onslide<2-5>{BOOL}
                             \onslide<2-6>{true:}%
                             \only<1>{\rlap{\sffamily
                                            \alert{Typed} target languages}}%
                             \only<7>{\rlap{\sffamily evaluate}}}%
                       \makebox[3em][l]{\bool\only<2-5>{\ value}}}};
        \draw [->] (lang) -- +(-80:3.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<7>{\hspace*{-16pt}compile}};
        \draw [->] (lang) -- +(-65:3cm)
            node [anchor=north west,inner xsep=0pt] {\only<7>{specialize}};
        \draw [->] (lang) -- +(-50:2.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<7>{\hspace*{-3pt}CPS transform}};
        \draw      (lang)    +(-35:2cm) node {\dots};
    \end{tikzpicture}

@@

    \strut
    \only<-3>{It should be obvious in the metalanguage that interpreting
    a well-typed source term yields a well-typed target term.}%
    \only<4>{The term should be well-typed (and closed), so pattern matching
    in the metalanguage should always \textbf{obviously} succeed.}%
    \only<5>{Previous solutions use (and motivate) fancier types:\\
    generalized abstract data types, dependent types.}%
    \only<6>{Our simple solution is to be \textbf{finally tagless:}\\
    replace term constructors by cogen functions.}%
    \only<7>{The term accommodates \textbf{multiple interpretations}\\
    by abstracting over the cogen functions and their types.}%
    \strut
\end{frame}
